---
title: 'Data For Online Quiz #1'
output: pdf_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 6

The following document will help you answer questions 6-10 of the online quiz. It will test your ability to use statistical terminology to describe the data presented.
I will read in some data and then we can look at some plots together.

The questions on the online quiz will ask you to provide a short answer response.

We will start by examining a grouped frequency distribution of *Anolis* lizards and their forelimb lengths.

```{r Question 6, echo=FALSE}
AnolisData <- read.csv("/Users/aconith/Library/CloudStorage/OneDrive-DePaulUniversity/DePaul/Teaching/2024SQ/Conith/GitHub/BIO206_BIOSTATS/Quiz_1/AnolisDataQuiz.csv")

hist(AnolisData$ForeLimbLength, breaks = 5, xlab = "Forelimb Length (cm)", ylab = "Frequency", main = expression('Q6, Histogram of '*italic(Anolis)*' Forelimb Lengths'))

```

## Question 7

We will now look over distribution of digit lengths in our sample.

```{r Question 7, echo=FALSE}
AnolisData <- read.csv("/Users/aconith/Library/CloudStorage/OneDrive-DePaulUniversity/DePaul/Teaching/2024SQ/Conith/GitHub/BIO206_BIOSTATS/Quiz_1/AnolisDataQuiz.csv")

hist(AnolisData$Digit5Length, breaks = 8, xlab = "Digit 5 Length (cm)", ylab = "Frequency", main = expression('Q7, Histogram of '*italic(Anolis)*' Digit 5 Lengths'))

```

## Question 8 and question 9

We will now look over distribution of limb and tail lengths in our sample.

In the short answer prompt on your online quiz describe the type and strength of the correlation (or lack thereof), and mention which measure is the independent variable and dependent variable for plot 8a.

```{r Question 8, echo=FALSE}
AnolisData <- read.csv("/Users/aconith/Library/CloudStorage/OneDrive-DePaulUniversity/DePaul/Teaching/2024SQ/Conith/GitHub/BIO206_BIOSTATS/Quiz_1/AnolisDataQuiz.csv")
par(mfrow=c(1,2), pty='s')
plot(AnolisData$BodyLength,AnolisData$ForeLimbLength, pch=19, cex.main=0.8, main = expression('Q8a, '*italic(Anolis)*' body length by forelimb length'), xlab="Body length (cm)", ylab="Forelimb Length (cm)", xlim = c(0,40),cex=0.6)
plot(AnolisData$BodyLength,AnolisData$TailLength, pch=19, cex.main=0.8, main = expression('Q8b, '*italic(Anolis)*' body length by tail length'), xlab="Body length (cm)", ylab="Tail Length (cm)", xlim = c(0,40),cex=0.6)

par(mfrow=c(1,1), pty='s')

```
## Question 10

We will now look over two different line plots.

In the short answer prompt on your online quiz describe what plots 9a and 9b depict. Also, for both plots list what the data type is for each variable (*i.e.*, continuous, discrete, categorical).

```{r Question 10, echo=FALSE}
AnolisData <- read.csv("/Users/aconith/Library/CloudStorage/OneDrive-DePaulUniversity/DePaul/Teaching/2024SQ/Conith/GitHub/BIO206_BIOSTATS/Quiz_1/AnolisDataQuiz.csv")

suppressWarnings(suppressMessages(library(ggplot2)))
suppressWarnings(suppressMessages(library(cowplot)))


x <- 1:17
y <- 5*exp(0.1*x)
Z<-cbind.data.frame(x,y)

A<-ggplot(data=Z,aes(x=x,y=y))+
       geom_line() +
  theme_bw() +
  xlab("Embryonic Stage") +
  ylab("Body Length (cm)") +
  ggtitle("Q9a, Body length through 17 stages") +
  theme(aspect.ratio=1,  plot.title = element_text(size = 10))

B<-ggplot(AnolisData, aes(x = BodyLength)) +
          geom_freqpoly(bins=10) +
  theme_bw() +
  xlab("Body Length (cm)") +
  ylab("Frequency") +
  ggtitle("Q9b, Frequency Distribution of Body Lengths") +
  theme(aspect.ratio=1, plot.title = element_text(size = 10))

plot_grid(A,B)

```

